---
title: "Social Network - Assignment #2"
author: "Ekin Kızıldaş - Gür Piren"
format:
  html:
    toc: true
    number-sections: true
    toc-depth: 3
date: "2025-05-14"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Link prediction Project

- Choose a network 0f 500 nodes and 5000+ edges
- Import and visualize the network
- Delete a fraction of real edges in the network and create a table
of those links deleted (positive class) and of links non-present
(negative class)
- Generate a number of proximity/similarty metrics heuristics for
each link in the positive and negative class
- Train a binary classifier to predict the links, i.e., to predict the
class (positive/negative) using those heuristics. Use cross-
validation.
- Evaluate the precision of the model. Which heuristic is the most
important. Why do you think it is the most important?
- Comment on potential ways to improve the link prediction

**Social Network Analysis Class Roadmap**

**Intro**

**1. Choose a network**

Use the network from homework 2 or select a new one (500+ nodes, 5000+ links, preferably social). Register choice on Aula Global.

```{r}

```


**2. Import network data**

Load data into R and create a graph using the igraph package. Use read.graph() or similar to import and graph_from_data_frame() to construct the graph.

```{r}

```


**3. Link Prediction Project**

**3.1.**

Delete a fraction of real edges: Randomly remove 10-20% of edges using delete_edges() in igraph. Create a table of deleted (positive) and non-existing (negative) links with sample() for negative class.

```{r}

```


**3.2.**

Generate proximity/similarity metrics: Compute heuristics (e.g., common neighbors, Jaccard, Adamic-Adar) for all links using similarity() functions in igraph or custom calculations.

```{r}

```


**3.3.**

Train a binary classifier: Use caret or randomForest in R to train a classifier (e.g., logistic regression, random forest) on heuristics with cross-validation (trainControl for k-fold CV).

```{r}

```


**3.4.**

Evaluate model precision: Assess model using confusionMatrix() in caret to calculate precision, recall, and AUC. Identify key heuristic via feature importance (varImp()).

```{r}

```


**3.5.**

Comment on improvements: Suggest incorporating node attributes or advanced algorithms (e.g., graph neural networks) to enhance prediction accuracy.


```{r}

```


